---
title: "PROGRAMMING"
author: "Jana Flakova"
date: "2025-06-18"
output:
  pdf_document: default
  html_document: default
---

## Part 2 - Data Sourcing

### 2.1 Load in the Data

We'll be using two datasets in this project:

1.  A custom dataset on the Gender Pay Gap in the EU (loaded from a CSV)
2.  A dataset on sectoral wages and gender workforce distribution in Germany (2023)

```{r Libraries, message=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r Datasets}
TableCountries = read.csv("data/TableCountries.csv")  
Germany_subgroup = read.csv("data/Germany_subgroup/Germany_subgroup.csv")
```

### 2.2 Short Summary

```{r Preview Datasets}
head(TableCountries)
head(Germany_subgroup)
```

### 2.3 Metadata Description

#### Main Dataset: EU Gender Pay Gap Data (2021–2023)

-   **Title**: EU Gender Pay Gap and Workforce Composition\
-   **Years Covered**: 2021, 2022, 2023\
-   **Geographic Scope**: All EU countries + EU-27 aggregate\
-   **Sources**: Eurostat, EIGE\
-   **Units**:
    -   `monthly_income`: Euros (gross, monthly)\
    -   `gender_pay_gap`: Percentage (%)\
    -   `p_female`, `p_male`: Share in workforce (0–1)
-   **Variables**:
    -   `country`: Country name (categorical)\
    -   `year`: Observation year (numeric)\
    -   `gender_pay_gap`: Unadjusted gender pay gap\
    -   `monthly_income`: Average monthly gross income\
    -   `p_female` / `p_male`: Workforce gender shares
-   **Collection Method**: Administrative statistics, not survey-based

------------------------------------------------------------------------

#### Subgroup Dataset: Germany by Sector (2023)

-   **Title**: Sectoral Gender Pay Gap and Earnings – Germany\
-   **Year**: 2023 (with income data likely from 2022)\
-   **Scope**: Sector-level data for Germany\
-   **Sources**: Statista, Destatis\
-   **Units**:
    -   `income_eur`: Euros (gross, monthly)\
    -   `gpg_2023`: Percentage (%)\
    -   `p_female`, `p_male`: Share in workforce (0–1)
-   **Variables**:
    -   `sector`: Economic sector name (categorical)\
    -   `income_eur`: Average monthly income\
    -   `gpg_2023`: Gender pay gap\
    -   `occupation_category`: Dominant occupational group\
    -   `p_female` / `p_male`: Workforce gender shares
-   **Collection Method**: Aggregated administrative data
